---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mcrutils

<!-- badges: start -->
[![Codecov test coverage](https://codecov.io/gh/mcaselli/mcrutils/graph/badge.svg)](https://app.codecov.io/gh/mcaselli/mcrutils)
[![R-CMD-check](https://github.com/mcaselli/mcrutils/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mcaselli/mcrutils/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of mcrutils is to ...

## Installation

You can install the development version of mcrutils from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("mcaselli/mcrutils")
```

## Examples


### Normalize logical columns

For data frames or tibbles that have character or factor columns storing 
logical data, as may happen when reading from a database, CSV, or Excel file,
use `normalize_logicals()` to find and convert these columns to logical type.
This is a nice one-liner in a `dplyr` pipe

```{r example}
library(mcrutils)

ugly_df <- data.frame(
  logical_char = c("T", "F", "T"),
  logical_factor = factor(c("TRUE", "FALSE", "TRUE")),
  non_logical_char = c("a", "b", "c"),
  non_logical_factor = factor(c("x", "y", "z")),
  mixed_char = c("T", "F", "a"),
  mixed_factor = factor(c("TRUE", "FALSE", "x")),
  numeric_col = c(1.1, 2.2, 3.3),
  stringsAsFactors = FALSE
)

df <- ugly_df |> normalize_logicals()

sapply(df, class)
```

### Year-to-date helpers

`mcrutils` provides a handful functions that can be helpful in creating 
year-to-date analyses

With `ytd_bounds()`, quickly find the date bounds of the latest year-to-date 
period in vector of dates, possibly spanning multiple years:
```{r}
c(
  "2023-01-01", "2023-06-15", "2023-12-31",
  "2024-01-01", "2024-03-15", "2024-07-15"
) |>
  as.Date() |>
  ytd_bounds()
```


`is_ytd_comparable()` is a logical vector that indicates whether the dates
in a vector are within a year-to-date period relative to a given `end_date`. 

This is useful as a filter criteria to create summaries of comparable periods 
in prior years.


```{r}
library(dplyr)

sales <- data.frame(
  date = seq(to= as.Date("2025-06-01"), 
            by = "month",
            length.out=30),
  amount = rpois(30, lambda = 100)
)

bounds <- ytd_bounds(sales$date)

sales |>
  filter(is_ytd_comparable(date, max(bounds))) |>
  group_by(year = lubridate::year(date)) |>
  summarise(ytd_sales = sum(amount))

```


With `py_dates()` you can rollback a vector of dates to the same period in the
previous year, moving any ficticious dates to the prior valid day.

```{r}
c("2024-01-01", "2024-02-29", "2025-07-15") |>
  as.Date() |>
  py_dates()
```
